MSc Project - Modelling a Double-Barred Galaxy as a Double Binary

### 2017-08-21 workspace/DoubleBars/MScProjectDoubleBarsv2.tex
Draft

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage[skip=0pt]{caption}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
%\setlength\abovecaptionskip{-5pt}
\author{Howard Kinsman}
\title{MSc Project - Modelling a Double-Barred Galaxy as a Double Binary}
\begin{document}
\maketitle
\begin{abstract}
TODO
\end{abstract}
\section{Introduction}
\subsection{Double-Barred Galaxies} 
Double-barred galaxies, where a smaller inner bar is nested inside a larger outer bar, were first observed in 1975 by de Vaucouleurs \citep{vaucouleurs}, however it was only in the 1990s that double-barred galaxies
were recognised as being a distinct category of galaxies. The current estimates of double-barred galaxy frequencies are $\approx30\%$ of barred galaxies and $\approx20\%$ of all galaxies \citep{erwin3}.
\cite{erwin2} surveyed 38 barred galaxies and found a total of 10 double-barred galaxies with no preference for Hubble type. They found that inner bars were about 12\% the size of the outer bars with
semi-major axes from $\approx250$pc to 1 kpc. Secondary bars probably rotate independently of, but in the same direction as their associated primary bars \cite{erwin2}.
\cite{erwin1} believe that inner bars probably rotate independently of their outer bars and suggest that they are relatively long-lived structures.
Inner bar size has been found to correlate slightly with outer bar size \citep{erwin3}. The median size ratio is $\approx0.12$ with an upper limit of $\approx0.25$ which is consistent with the theory that
inner bars cannot be too large without disrupting the orbit of the outer bar.

@@

It has been suggested that nested bars may provide a mechanism for fueling active galactic nuclei \citep{shlosman} however more recently \citep{erwin3} believe that inner bars only play a minor role in this 
activity. \citep{erwin1} compared single and double-barred galaxies and found that there was no significant difference in nuclear activity. To investigate this theory further \cite{lorenzo1} carried out a 
study of NGC 357. This study aimed to present the first detailed morphological, kinematical and stellar population analysis of the bulge, inner and outer bars of a double-barred galaxy. They concluded that
the bulge and inner bar show similar stellar population properties whilst the outer bar was less metal-rich. This result implied that, for the case of NGC 357, the outer bar was formed in a shorter timescale than
the inner structures and so disagrees with traditional idea that gas flown along the outer structure triggers star formation causing the formation of the inner structure. This was strengthened by a further study of
four double-barred galaxies by \cite{lorenzo2} where they concluded that the stellar populations of inner bars are younger and more metal-rich than the outer bars. This suggests that at present inner bars play a
moderate or even minor role in the morphological evolution of double-barred galaxies \citep{lorenzo2}.

@@

Methods for the study of N-body simulations have changed considerably since the advent of computers. In a study of N-body simulations \cite{trenti} mentions a pioneering attempt by Holmberg in 1941 which used 
light bulbs and galvanometers to track the evolution of 37 particles. However, it wasn't until the early seventies and the advent of computers that this field really got started. 
Nowadays simulations can up to a million particles \citep{heggie} which is of particular relevance to the study of globular
clusters. I refer the reader to \cite{aarseth} for much more detailed explanation of the field of N-body simulations.

@@

In an attempt to explain the dynamics of the orbits of nested bars \cite{macie7} introduced the concept of `loops'. They wanted to investigate how two independently rotating bars can maintain
themselves as gravitating systems. A gravitational potential consisting of two concentric bisymmetric bars that are independently rotating is not stationary in any reference frame \cite{macie7}. In the potential
of a single bar particles would stay on a stable closed periodic orbit but with nested bars orbits will not be closed because the particles would be experience two separate forces. The `loop' therefore was
introduced as an extension of the concept of orbit. Particles in a `loop' would return to their original positions every time the bars come back to the same relative orientation \citep{macie7}. This lead to
the conclusion that loops supporting the inner bar are thicker with parallel bars and thinner with perpendicular bars \citep{macie3}. This also suggets that the two bars do not rotate through each other as rigid bodies
but the inner bar should pulsate and that the bars should spend more time nearly orthognal\citep{macie3}. These predictions were confirmed by \citep{debattista} and \citep{shen1}. 
This was further confirmed by \cite{macie5} who found
that the inner bar should end well within its corotation radius and that the inner bar extends further out for lower pattern speeds than higher ones; the inner bar pulsates; and that the angular velocity of the inner
bar is not uniform and that faster inner bars rotate more coherently.

\cite{wozniak} also carried out double-barred galaxy simulations and achieved long-lived inner bars lasting for up to 7 Gyr. These were N-body/hydrodynamical simulations. They found that the two bar lengths, ratio of
pattern speeds, as well as the age of the inner stellar bar populations in the simulations matched well with observations. They concluded that because, unlike previous simulations, there was no overlap between primary bar 
inner Lindblad resonance (ILR) and the inner bar corotation (CR) or any kind of resonance overlap, then some other kind of mechanism must feed the central waves \citep{wozniak}. It is suggested that star 
formation may be responsible for bringing energy into the inner bar.

\cite{garzon} studied the dynamical evolution of two isolated bars, within the same galaxy, under their mutual gravitational interaction. They considered two cases purely analytically: rotation of rigid bodies and
rotation of deformable rotation. They found that the bars oscillate in size according to the relative angle between the bars. This is broadly in agreement with \citep{debattista}.
What follows is an attempt to verify the calculations in \citep{garzon} but using a simpler model of a double-barred galaxy using a double binary.

@@

I then proceeded to perturb the system by gradually increasing the mass of the inner binary. As the system became unstable I compensated for
the increasing mass of the inner binary by increasing the velocity of the outer binary and also decreasing the velocity and increasing the binary 

@@

Configuration 32 was by far the longest running simulation showing almost three complete orbits of the outer binary and many orbits of the inner binary. It showed no signs of destabilizing and the 
simulation only ended because of a `Step size underflow' in the integrator. This error occurs when the adaptive stepsize becomes too small and the integrator is effectively not making any
progress with the integration and so ends the simulation. The data shows that as the periodic ratios tends towards 1:21 the system considerably stabilizes. 1:21 could therefore possibly be an orbital resonance. 
Attempts at modifying the initial conditions of configuration 32 even slightly resulted in chaotic systems.

@@

%\citep{garzon}, \citep{erwin1}, \citep{macie1}, \citep{macie2}, \citep{macie3}, \citep{macie4}, \citep{macie5}
%\citep{macie6}, \citep{macie7}, \citep{manos}, \citep{shen1}, \citep{shen2}, \citep{debattista}, \citep{malhotra}
 %\citep{du2} \citep{manos}, \citep{shen2}

@@

%\bibitem[Du et. al.(2016)]{du2}Du, M., Debattista, V., Shen, J., Kimematic Properties of Double-Barred Galaxies: Simulations vs. Integral-Field Observations, 2016, arXiv:1607.00585v1

@@

\bibitem[Garzon(2014)]{garzon}Garzon, F., Lopez-Corredoira, M., Dynamical evolution of two associated galactic bars, 2014, arXiv:1409.1916.v1

@@

%\bibitem[Manos and Athanassoula(2011)]{manos}Manos, T., Athanassoula, E., Regular and chaotic orbits in barred galaxies - I. Applying the SALI/GALI method to explore their distribution in several models, 2011, arXiv:1102.1157v2

>>>>

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage[skip=0pt]{caption}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{MSc Project - Modelling a Double-Barred Galaxy as a Double Binary}
\begin{document}
\maketitle
\begin{abstract}
TODO
\end{abstract}
\section{Introduction}
\subsection{Double-Barred Galaxies} 
Double-barred galaxies, where a smaller inner bar is nested inside a larger outer bar, were first observed in 1975 by de Vaucouleurs \citep{vaucouleurs}, however it was only in the 1990s that double-barred galaxies
were recognised as being a distinct category of galaxies. The current estimates of double-barred galaxy frequencies are $\approx30\%$ of barred galaxies and $\approx20\%$ of all galaxies \citep{erwin3}.
\cite{erwin2} surveyed 38 barred galaxies and found a total of 10 double-barred galaxies with no preference for Hubble type. He found that inner bars were about 12\% the size of the outer bars with
semi-major axes from $\approx250$pc to 1 kpc. Secondary bars probably rotate independently of, but in the same direction as their associated primary bars \citep{erwin2}.
\cite{erwin1} believe that inner bars probably rotate independently of their outer bars and suggest that they are relatively long-lived structures.
Inner bar size has been found to correlate slightly with outer bar size \citep{erwin3}. The median size ratio is $\approx0.12$ with an upper limit of $\approx0.25$ which is consistent with the theory that
inner bars cannot be too large without disrupting the orbit of the outer bar.

@@

It has been suggested that nested bars may provide a mechanism for fueling active galactic nuclei \citep{shlosman} however more recently, \citep{erwin3}, it was found that inner bars only play a minor role in this 
activity. \cite{erwin1} compared single and double-barred galaxies and found that there was no significant difference in nuclear activity. To investigate this theory further \cite{lorenzo1} carried out a 
study of NGC 357. This study aimed to present the first detailed morphological, kinematical and stellar population analysis of the bulge, inner and outer bars of a double-barred galaxy. They concluded that
the bulge and inner bar show similar stellar population properties whilst the outer bar was less metal-rich. This result implied that, for the case of NGC 357, the outer bar was formed in a shorter timescale than
the inner structures and so disagrees with traditional idea that gas flown along the outer structure triggers star formation causing the formation of the inner structure. This was strengthened by a further study of
four double-barred galaxies by \cite{lorenzo2} where they concluded that the stellar populations of inner bars are younger and more metal-rich than the outer bars. This suggests that at present inner bars play a
moderate or even minor role in the morphological evolution of double-barred galaxies \citep{lorenzo2}.

@@

Methods for the study of N-body simulations have changed considerably over the years. In a study of N-body simulations \cite{trenti} mentions a pioneering attempt by Holmberg in 1941 which used 
light bulbs and galvanometers to track the evolution of 37 particles. However, it wasn't until the early seventies and the advent of computers that this field really got started. 
Nowadays simulations can up to a million particles \citep{heggie} which is of particular relevance to the study of globular
clusters. I refer the reader to \cite{aarseth} for much more detailed explanation of the field of N-body simulations.

@@

In an attempt to explain the dynamics of the orbits of nested bars \cite{macie7} introduced the concept of `loops'. They wanted to investigate how two independently rotating bars can maintain
themselves as gravitating systems. A gravitational potential consisting of two concentric bisymmetric bars that are independently rotating is not stationary in any reference frame \citep{macie7}. In the potential
of a single bar particles would stay on a stable closed periodic orbit but with nested bars orbits will not be closed because the particles would be experience two separate forces. The `loop' therefore was
introduced as an extension of the concept of orbit. Particles in a `loop' would return to their original positions every time the bars come back to the same relative orientation \citep{macie7}. This lead to
the conclusion that loops supporting the inner bar are thicker with parallel bars and thinner with perpendicular bars \citep{macie3}. This also suggets that the two bars do not rotate through each other as rigid bodies
but the inner bar should pulsate and that the bars should spend more time nearly orthognal\citep{macie3}. These predictions were confirmed by \citep{debattista} and \citep{shen1}. 
This was further confirmed by \cite{macie5} who found
that the inner bar should end well within its corotation radius and that the inner bar extends further out for lower pattern speeds than higher ones; the inner bar pulsates; and that the angular velocity of the inner
bar is not uniform and that faster inner bars rotate more coherently.

\cite{wozniak} also carried out double-barred galaxy simulations and achieved long-lived inner bars lasting for up to 7 Gyr. These were N-body/hydrodynamical simulations. He found that the two bar lengths, ratio of
pattern speeds, as well as the age of the inner stellar bar populations in the simulations matched well with observations. He concluded that because, unlike previous simulations, there was no overlap between primary bar 
inner Lindblad resonance (ILR) and the inner bar corotation (CR) or any kind of resonance overlap, then some other kind of mechanism must feed the central waves \citep{wozniak}. It is suggested that star 
formation may be responsible for bringing energy into the inner bar.

\cite{garzon} studied the dynamical evolution of two isolated bars, within the same galaxy, under their mutual gravitational interaction. They considered two cases purely analytically: rotation of rigid bodies and
rotation of deformable rotation. They found that the bars oscillate in size according to the relative angle between the bars. This is broadly in agreement with \citep{debattista}.
What follows is an attempt to verify the calculations in \cite{garzon} but using a simpler model of a double-barred galaxy using a double binary.

@@

The system was then perturbed by gradually increasing the mass of the inner binary. As the system became unstable the increasing mass of the inner binary 
was compensated for by increasing the velocity of the outer binary and also decreasing the velocity and increasing the binary 

@@

Configuration 32 was by far the longest running simulation showing almost three complete orbits of the outer binary and many orbits of the inner binary. It showed no signs of destabilizing and the 
simulation only ended because of a `Step size underflow' in the integrator. This error occurs when the adaptive stepsize becomes too small and the integrator is effectively not making any
progress with the integration and so ends the simulation. The data shows that as the periodic ratios tends towards 1:21 the system considerably stabilizes. 1:21 could therefore possibly be an orbital resonance. 
Attempts at modifying the initial conditions of configuration 32 even slightly resulted in chaotic systems. The concept of a resonance is very familiar to us all: pushing someone on a playground swing at its natural
frequency will rapidly increase the swing amplitude \citep{malhotra}. Resonances occur naturally in the Solar system and there are three main types of resonance: spin-orbit, secular and mean motion \citep{malhotra}.
The most obvious example of a spin-orbit resonance is the Earth/Moon system where the Moon is spin-locked and only one side of the Moon is visible from the Earth. Secular resonances occur where the precession of two
orbits is synchronised. However, the 1:21 resonance, is an example of a mean-motion resonance where the orbital period of two bodies are close to a ratio of small integer \citep{malhotra}. Orbital resonances
generally results in destabilizing a system however occasionally they may be stable and remain in resonance.
In the standard model of double-barred galaxies \citep{macie7} it is assumed that resonant coupling between bars is present: CR of the inner bar coincides with the ILR of the outer bar. Although \cite{macie4} 
found that resonant coupling between the bars may not be the dominant factor reducing chaos in the system.

@@

@@

@@

\bibitem[Garzon and Lopez-Corredoira(2014)]{garzon}Garzon, F., Lopez-Corredoira, M., Dynamical evolution of two associated galactic bars, 2014, arXiv:1409.1916.v1

@@
